---
title: "BayesVolcano"
package: BayesVolcano
author: "Katja Danielzik (katja.danielzik@uni-due.de)"

output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BayesVolcano}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Data requirements

BayesVolcano requires as input:

- posterior samples of a Bayesian model fit stored in a data frame, where columns are parameters and rows are iterations. 
A wrapper function (extract_fit()) to extract the posterior samples for fits generated by [rstan](https://doi.org/10.32614/CRAN.package.rstan) or [brms](https://doi.org/10.32614/CRAN.package.brms) is included in the package.
The usage of extract_fit() requires the installation of rstan or brms.

```{r setup}
library(BayesVolcano)
```
```{r}
data("posterior")
head(posterior[, 1:4])
```

- a data frame mapping parameter names to biological entities with at least a column named "parameter" and a column named "label". Additional columns can be provided that might be helpful for later visualization (can be categorical (here: "group") or numerical (here: "value"))

```{r}
data("annotation_df")
head(annotation_df)
```

# Background
Bayesian models are used to estimate effect sizes (e.g., gene expression changes,
metabolite or protein abundance differences, drug response effects) while accounting for uncertainty, 
small sample sizes, and complex experimental designs.
However, Bayesian posteriors of models with many parameters are often difficult to interpret at a glance.

One way to quickly identify important biological changes based on frequentist analysis 
are volcano plots (using fold-changes and p-values).

Bayesian volcano plots bring together the explicit treatment of uncertainty in
Bayesian models and the familiar visualization of volcano plots.

We are not the first to think about the concept of Bayesian volcano plots: [Sousa et al. 2020](https://doi.org/10.1016/j.aca.2019.11.006) introduced them as a single
use case (pi-value= 1 - b-value)
but to our knowledge we are the first to provide an R-package
for easy calculation of pi-values and visualization. 

# Workflow
## Preparation of plotting dataframe

The prepare_volcano_df() function takes as input the posterior and the annotation
data frame. For **every parameter listed in the annotation data frame** the function
calculates one pi-value on basis of the following formula:

With *i* being one entity that was modeled, *param* the estimated parameter and *t*
the central parameter value corresponding to no effect (default t=0), the pi-
value is calculated as:

![equation](https://latex.codecogs.com/svg.image?%5Cpi_%7Bi%7D=2*max%5Cleft(%5Cint_%7Bparam_%7Bi%7D=-%5Cinfty%7D%5E%7Bt%7Dp(param_%7Bi%7D)%5C,dparam_%7Bi%7D,%5Cint_%7Bparam_%7Bi%7D=t%7D%5E%7B%5Cinfty%7Dp(param_%7Bi%7D)%5C,dparam_%7Bi%7D%5Cright)-1)

In other words, the pi-value answers the question how many of the posterior samples
lie beyond the central parameter value corresponding to no effect. The pi-value
can range from 0 (posterior samples are equally distributed on both sides of 
the central parameter value) to 1 (all posterior samples are below or above
the central parameter value).

The user can specify the central parameter value corresponding to no effect 
(in the package referred to as **"zero effect"**) and further can specify which
credible interval (CrI) should be calculated from the posterior. This can later be visualized in the volcano plot.

```{r}
library(BayesVolcano)
results <- prepare_volcano_df(
  posterior = posterior,
  annotation_df = annotation_df,
  zero.effect = 0.5, # for demonstration purposes, usually 0 is a valid choice
  CrI.low = 0.025, # 2.5%
  CrI.high = 0.975 # 97.5%, in combination with CrI.low translates to 95% CrI
)
```

This returns a list with two elements:

- a data frame with the columns parameter, pi.value, parameter.median, parameter.low (CrI.low),
parameter.high (CrI.high), label (from annotation data frame) and additional
columns (here "group" and "value") from the annotation data frame

```{r}
head(results$result)
```

- the meta information (user defined zero effect, CrI.low, and CrI.high)

```{r}
results$meta
```

## Volcano Plot
The plot_volcano() function is a wrapper function for convenient plotting
of the results$results data frame with [ggplot2](https://ggplot2.tidyverse.org/)
that automatically includes the meta information. It returns a ggplot object that
can be further customized by the user.

The simplest version is: 

```{r,fig.width=7,fig.height=6}
plot_volcano(results)
```

Where each point corresponds to one modeled entity. The
set zero effect for pi is displayed as vertical black line and described in the
subtitle.

Titles and x-axis labels can be set with:
```{r,fig.width=7,fig.height=6}
plot_volcano(results,
  title = "My amazing plot",
  xlab = "My informative parameter"
)
```


To add the set CrI use:
```{r,fig.width=7,fig.height=6}
plot_volcano(results,
  CrI = TRUE
)
```

This adds the credible intervals as grey lines and adds a description of the errorbar
to the subtitle.

Another possibility is to display the CrI width (distance between lower and 
upper boundary) as point size. For this set CrI.width = TRUE. Below I display
the CrI and the summary as point size in one plot.

```{r,fig.width=7,fig.height=6}
plot_volcano(results,
  CrI = TRUE,
  CrI.width = TRUE
)
```

Wide CrI translate into small points, narrow CrI into big points.

To color code the points the user can choose a character of numerical column
from the result data frame (annotation_df).

```{r,fig.width=7,fig.height=6}
plot_volcano(results,
  color = "group"
)

plot_volcano(results,
  color = "value"
)
```

To add labels the [ggrepel](https://doi.org/10.32614/CRAN.package.ggrepel)
package is employed:

```{r,fig.width=7,fig.height=6}
plot_volcano(results,
  label = "label"
)
```

To only display labels above thresholds for the parameter value and pi set:

```{r,fig.width=7,fig.height=6}
plot_volcano(results,
  label = "label",
  label.parameter.threshold = 1.5, # parameter threshold
  label.pi.threshold = 0.95
) # pi threshold
```

Grey lines are added to visualize the threshold and a subtitle is added.

Of course color, labels and CrI width can be combined:

```{r,fig.width=7,fig.height=6}
plot_volcano(results,
  CrI.width = TRUE,
  color = "group",
  label = "label",
  label.parameter.threshold = 1.5,
  label.pi.threshold = 0.95,
  title = "My plot",
  xlab = "My informative parameter"
)
```

# References
Julie de Sousa, Ondřej Vencálek, Karel Hron, Jan Václavík, David Friedecký, Tomáš Adam,
Bayesian multiple hypotheses testing in compositional analysis of untargeted metabolomic data,
Analytica Chimica Acta, Volume 1097, 2020, Pages 49-61, ISSN 0003-2670,
https://doi.org/10.1016/j.aca.2019.11.006

Corresponding GitHub Repository: https://github.com/sousaju/BayesVolcano
