---
title: "BayesVolcano"
package: BayesVolcano
author: "Katja Danielzik (katja.danielzik@uni-due.de)"

output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BayesVolcano}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Data requirements

BayesVolcano requires as input:

- posteriors of a Bayesian model fit stored in a data frame, where columns are parameters and rows are iterations. A wrapper function to extract the posterior exists for fits generated by [rstan](https://mc-stan.org/rstan) or [brms](https://paulbuerkner.com/brms) with the helper function extract_fit().
The usage of extract_fit() requires installation of rstan or brms.

```{r setup}
library(BayesVolcano)
```
```{r}
data("posterior")
head(posterior[,1:4])
```

- a data frame mapping parameter names to biological entities with at least a column named "parameter" and a column named "label". Additional columns can be provided that might be helpful for later visualization (can be categorical (here: group) or numerical (here:value))

```{r}
data("annotation_df")
head(annotation_df)
```

# Background
Bayesian models are used to estimate effect sizes (e.g., gene expression changes,
protein abundance differences, drug response effects) while accounting for uncertainty, 
small sample sizes, and complex experimental designs.
However, Bayesian posteriors of models with many parameters are often difficult to interpret at a glance.

One way to quickly identify important biological changes based on frequentist analysis 
are volcano plots (using fold-changes and p-values).

Bayesian volcano plots bring together the uncertainty aware power of Bayesian 
models and the familiar visualization of volcano plots.

We are not the first to think about the concept of Bayesian volcano plots [Sousa et al. 2020](https://doi.org/10.1016/j.aca.2019.11.006) introduced them as a single
use case (their b-values correspond to our pi-values)
but to our knowledge we are the first to provide an R-package
for easy calculation of pi-values and visualization. 

# Workflow
## Preparation of plotting dataframe

The prepare_volcano_df() function takes as input the posterior and the annotation
data frame. For every parameter listed in the annotation data frame the function
calculates one pi-value on basis of the following formula:

With *i* being one entity that was modeled, *param* the estimated parameter
and *t* the threshold for which to calculate the pi-value.

![equation](https://latex.codecogs.com/svg.image?%5Cpi_%7Bi%7D=2*max%5Cleft(%5Cint_%7Bparam_%7Bi%7D=-%5Cinfty%7D%5E%7Bt%7Dp(param_%7Bi%7D)%5C,dparam_%7Bi%7D,%5Cint_%7Bparam_%7Bi%7D=t%7D%5E%7B%5Cinfty%7Dp(param_%7Bi%7D)%5C,dparam_%7Bi%7D%5Cright)-1)

The user further can specify which credible interval (CrI) should be calculated
from the posterior. This can later be visualized in the volcano plot.

```{r}
library(BayesVolcano)
results <- prepare_volcano_df(posterior=posterior, 
                   annotation_df = annotation_df, 
                   threshold = 0.5, # for demonstration purposes, with centered parameters usually 0 is a valid choice
                   CrI.low = 0.025, #2.5%
                   CrI.high = 0.975) # 97.5%, in combination with CrI.low translates to 95% CrI

```

This returns a list with two elements:
- a data frame with the parameter, pi.value, parameter.median, parameter.low (CrI.low),
parameter.high (CrI.high), label (from annotation data frame) and additional
columns (here "group") from the annotation data frame

```{r}
head(results$result)
```
- the meta information (set threshold, CrI.low, and CrI.high)
```{r}
results$meta
```

## Volcano Plot
The plot_volcano() function is a wrapper function for convenient plotting
of the results$results data frame with [ggplot2](https://ggplot2.tidyverse.org/)
that automatically includes the
meta information. It returns a ggplot object that can be further customized by
the user.

The simplest version is: 

```{r,fig.width=7,fig.height=6}
plot_volcano(results)
```

Where each point corresponds to one parameter of the Bayesian posterior. The
set threshold for pi is displayed as vertical black line and described in the
subtitle.

Titles and x-axis labels can be set with:
```{r,fig.width=7,fig.height=6}
plot_volcano(results, 
             title = "My amazing plot",
             xlab = "My informative parameter")
```


To add the set CrI use:
```{r,fig.width=7,fig.height=6}
plot_volcano(results,
             CrI = TRUE)
```

This adds the credible intervals as grey lines and adds a description of the errorbar
to the subtitle.

To color code the points the user can choose a character of numerical column
from the annotation_df

```{r,fig.width=7,fig.height=6}
plot_volcano(results,
             CrI = FALSE,
             color = "group")

plot_volcano(results,
             CrI = FALSE,
             color = "value")
```

To add labels the [ggrepel](https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html)
is employed:

```{r,fig.width=7,fig.height=6}
plot_volcano(results, label = "label")
```

To only display labels above thresholds for the parameter value and pi set:

```{r,fig.width=7,fig.height=6}
plot_volcano(results, 
             label = "label",
             label.parameter.threshold = 1.5,
             label.pi.threshold = 0.95)
```

Grey lines are added to visualize the threshold and a subtitle is added.

Of course color and labels can be combined:
```{r,fig.width=7,fig.height=6}
plot_volcano(results, 
             color = "group",
             label = "label",
             label.parameter.threshold = 1.5,
             label.pi.threshold = 0.95,
             title = "My plot",
             xlab = "My informative parameter")
```

# References
Julie de Sousa, Ondřej Vencálek, Karel Hron, Jan Václavík, David Friedecký, Tomáš Adam,
Bayesian multiple hypotheses testing in compositional analysis of untargeted metabolomic data,
Analytica Chimica Acta, Volume 1097, 2020, Pages 49-61, ISSN 0003-2670,
https://doi.org/10.1016/j.aca.2019.11.006.
(https://www.sciencedirect.com/science/article/pii/S0003267019313492)

Corresponding GitHub Repository: https://github.com/sousaju/BayesVolcano